---
title: "Practical Meachine Learning - JHS"
author: "Arghya Ghosh"
date: "22 December 2015"
output: html_document
---

Background

Devices such as Jawbone Up, Nike FuelBand, and Fitbit now can collect large amount of data about personal activity relatively inexpensively. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. Other variables would be used to predict classe. The following report describes the steps taken to build classification model, use of cross validation and notes on the interpretation of the results.

Classe variable consists of the following:

Class A: exactly according to the specification
Class B: throwing the elbows to the front
Class C: lifting the dumbbell only halfway
Class D: lowering the dumbbell only halfway
Class E: throwing the hips to the front

More information is available in website: http://groupware.les.inf.puc-rio.br/har

The Write Up is divided into following sections

Objective of the exercise
Data Import
Exploratory Data Analysis
Model Building/ Evaluation
Cross Validation
Model Results & Interpretation


```{r}
cbind(Summary=summary(cars), SD_Speed=sd(cars$speed), SD_dist=sd(cars$dist))
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
hist(cars$speed)
hist(cars$dist)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
